## Organising the UNICEF download {#c01-organise}

When several MICS surveys are downloaded together, UNICEF packages them in a bulk zip of survey folders, further archives and wrapper directories. The data are usable, but unpacking many surveys by hand is tedious and invites path and naming errors.

AFLEARN provides one script, [`run-it.R`](https://github.com/datafirst-aflearn/MICS6-Guide/blob/main/Script/run-it.R), from its Script suite [attach a link] that unpacks the download and then runs the two harmonisation scripts (more on this in the later .

### What `run-it.R` does

In this order, it:

1.  sources [`prepare-mics-fs.R`](https://github.com/datafirst-aflearn/MICS6-Guide/blob/main/Script/prepare-mics-fs.R) and calls `prepare_mics_data()`;
2.  sources `harmonize-mics-fs-v1.0.R` ;
3.  sources `harmonize-mics-reading-v1.1.R` .

The unpack step is base R.

`prepare_mics_data()`:

- requires a `Script/` folder in the working directory, so a wrong working directory fails immediately;
- creates `Data/`, `Data/UNICEF/` and `Data/AFLEARN Harmonised Data/` if they are missing;
- searches for `MICS_Datasets.zip` (or `MICS_Dataset.zip`) in this order: the `zip=` argument, the project folder, `Data/`.
- extracts nested archives, keeps the survey `.sav` files and readme documentation, and removes wrapper folders;
- writes one clean folder per survey under `Data/UNICEF/`.

The bulk zip is kept by default.

### Before and after

A bulk UNICEF download may look like this:

``` text
MICS_Datasets.zip
└── survey folder
    └── another archive
        └── wrapper folder
            ├── fs.sav
            ├── hh.sav
            ├── hl.sav
            └── ...
```

After a successful run:

``` text
MICS project/
├── Data/
│   ├── UNICEF/
│   │   ├── BEN_2021_MICS6_v01_M/
│   │   │   ├── bh.sav
│   │   │   ├── ch.sav
│   │   │   ├── fs.sav
│   │   │   ├── hh.sav
│   │   │   ├── hl.sav
│   │   │   ├── wm.sav
│   │   │   └── ...
│   │   ├── GHA_2017_MICS6_v01_M/
│   │   │   └── ...
│   │   └── ...
│   └── AFLEARN Harmonised Data/
│       ├── mics6-fs-harmonised.dta
│       ├── mics6_fs_variable_crosswalk.xlsx
│       ├── mics6-reading-harmonised.dta
│      
└── Script/
    ├── run-it.R
    ├── prepare-mics-fs.R
    ├── harmonize-mics-fs-v1.0.R
    └── harmonize-mics-reading-v1.1.R
```

| Folder | Contents |
|------------------------------------|------------------------------------|
| `Data/UNICEF/` | Original country files, one folder per survey |
| `Data/AFLEARN Harmonised Data/` | Created empty, then filled by the two harmonisation scripts |
| `Script/` | Every R and Stata script you run; supplied by the AFLEARN download |

Each survey then has a predictable path such as `Data/UNICEF/BEN_2021_MICS6_v01_M/fs.sav`. The later scripts expect that structure.

### How to run it

1.  Make a folder for the project, for example `MICS project`.
2.  Download the `Script` folder from the [AFLEARN MICS6-Guide repository](https://github.com/datafirst-aflearn/MICS6-Guide) and unzip it into that folder, so you have `MICS project/Script`.
3.  Put `MICS_Datasets.zip` in the project folder.
4.  In R, make the project folder the working directory and source `run-it.R`:

```{r eval=FALSE}
setwd("C:/path/to/MICS project")
source("Script/run-it.R")
```

The working directory must be the project folder itself, the one holding `Script/`. If it is not, the script stops before creating anything.

To delete the bulk zip after a successful extraction, call the unzip step yourself:

```{r eval=FALSE}
source("Script/prepare-mics-fs.R")

prepare_mics_data(remove_zip = TRUE)
```

If the zip is somewhere else:

```{r eval=FALSE}
source("Script/prepare-mics-fs.R")

prepare_mics_data(
  zip = "C:/path/to/MICS_Datasets.zip"
)
```

Chapter 3 explains the files those scripts write and how to re-run one step. You do not need to source the harmonisation scripts separately on the first run.

### Check the result

Confirm that R can see the survey folders and files:

```{r eval=FALSE}
dir.exists("Data/UNICEF")

list.files("Data/UNICEF")

file.exists(
  "Data/UNICEF/BEN_2021_MICS6_v01_M/fs.sav"
)
```

Open one of the SPSS files:

```{r eval=FALSE}
library(haven)

fs <- read_sav(
  "Data/UNICEF/BEN_2021_MICS6_v01_M/fs.sav"
)

dim(fs)
```

::: alternativebox-header
**Keep the original download**
:::

::: alternativebox
Keep the original UNICEF zip in a separate raw-data location, even if you use AFLEARN's clean folder structure for analysis. Raw source data should remain unchanged so that the preparation process is reproducible.
:::

[What's in MICS6 data?](#c02-whats-in-data) explains the file structure and the FLS module. If you already know that structure, continue to [Preparing UNICEF data](#c03-prepare-unicef) for merges, the variables the scripts produced, and the reading construction.
