# Organising the UNICEF download {#c03-organise}

When several MICS surveys are downloaded together, UNICEF packages them in a bulk zip containing survey folders, further archives and wrapper directories.

The data are usable, but manually unpacking many surveys is tedious and creates opportunities for path and naming errors.

AFLEARN provides an R script, [`prepare-mics-fs.R`](https://github.com/datafirst-aflearn/MICS6-Guide/blob/main/Script/prepare-mics-fs.R), to do this automatically.

## What the script does

The script [`prepare-mics-fs.R`](https://github.com/datafirst-aflearn/MICS6-Guide/blob/main/Script/prepare-mics-fs.R):

- creates the working folders used throughout this guide;
- finds the UNICEF `MICS_Datasets.zip`;
- extracts the nested archives;
- retains the survey `.sav` files and readme documentation;
- removes unnecessary wrapper folders;
- creates one clean folder for each survey; and
- copies itself, and any AFLEARN harmonisation scripts it finds alongside the zip, into `Script/`.

It uses base R only, so no packages need to be installed first.

## Before

A bulk UNICEF download may look conceptually like this:

```text
MICS_Datasets.zip
└── survey folder
    └── another archive
        └── wrapper folder
            ├── fs.sav
            ├── hh.sav
            ├── hl.sav
            └── ...
```

## After

AFLEARN's script produces:

```text
Data/
├── UNICEF/
│   ├── BEN_2021_MICS6_v01_M/
│   │   ├── bh.sav
│   │   ├── ch.sav
│   │   ├── fs.sav
│   │   ├── hh.sav
│   │   ├── hl.sav
│   │   ├── wm.sav
│   │   └── ...
│   ├── GHA_2017_MICS6_v01_M/
│   │   └── ...
│   └── ...
├── IPUMS/
└── AFLEARN Harmonised Data/
Script/
└── prepare-mics-fs.R
```

The three folders under `Data/` separate the three kinds of data this guide uses:

| Folder | Contents |
|---|---|
| `Data/UNICEF/` | The original country files, one folder per survey, written by this script |
| `Data/IPUMS/` | Your own IPUMS MICS extracts. Created empty; see [Getting data from IPUMS MICS](#c01-ipums) |
| `Data/AFLEARN Harmonised Data/` | Created empty. The harmonisation scripts write their output here |

`Script/` sits alongside `Data/` and holds every R and Stata script you run.

Each survey therefore has a predictable path such as:

```text
Data/UNICEF/BEN_2021_MICS6_v01_M/fs.sav
```

That predictable structure is what the later harmonisation scripts expect.

## How to run it

Download [`prepare-mics-fs.R`](https://github.com/datafirst-aflearn/MICS6-Guide/blob/main/Script/prepare-mics-fs.R) from the [AFLEARN MICS6-Guide repository](https://github.com/datafirst-aflearn/MICS6-Guide) and place it next to `MICS_Datasets.zip`.

Then, in R:

```{r eval=FALSE}
setwd("C:/path/to/folder/with/zip/and/script")
source("prepare-mics-fs.R")
```

Sourcing the script runs it. It creates `Data/UNICEF/`, `Data/IPUMS/`, `Data/AFLEARN Harmonised Data/` and `Script/` under that folder, then fills in the UNICEF survey folders. Running it again is safe: folders that already exist are left alone.

If you are working inside the MICS6-Guide project, the script finds the project root and builds the same structure there, whichever subfolder you run it from.

The original bulk zip is retained by default.

If you deliberately want to delete it after successful extraction:

```{r eval=FALSE}
prepare_mics_data(remove_zip = TRUE)
```

If the zip is somewhere else:

```{r eval=FALSE}
source("prepare-mics-fs.R")

prepare_mics_data(
  zip = "C:/path/to/MICS_Datasets.zip"
)
```

## Check the result

Before continuing, check that R can see the survey folders and files:

```{r eval=FALSE}
dir.exists("Data/UNICEF")

list.files("Data/UNICEF")

file.exists(
  "Data/UNICEF/BEN_2021_MICS6_v01_M/fs.sav"
)
```

You can also confirm that R can open one of the SPSS files:

```{r eval=FALSE}
library(haven)

fs <- read_sav(
  "Data/UNICEF/BEN_2021_MICS6_v01_M/fs.sav"
)

dim(fs)
```

::: {.alternativebox-header}
**Keep the original download**
:::

::: {.alternativebox}
We recommend keeping the original UNICEF zip in a separate raw-data location, even if you use AFLEARN's clean folder structure for analysis. Raw source data should remain unchanged so that the preparation process is reproducible.
:::

## Next

If you need variables from more than one MICS file within a country, continue to [Merging MICS6 files within a country](#c03-merge).

If all the variables you need are already in `fs`, you can move directly to [Harmonising variables across countries](#c03-harmonise).
