## Merging MICS6 files within a country {#c03-merge}

MICS data are split across files because the files have different units of observation. You will often need to add information from another file to the selected-child (`fs`) data.

::: {.alternativebox-solo}
**Decide what one row of the final dataset represents before you merge anything.**
:::

For foundational-learning research, one row will usually represent the selected child in `fs`. Start with `fs` and add the variables you need.

UNICEF's MICS6 statistical analysis manual includes guidance on merging MICS datasets and describes the identifiers used across questionnaire files. See the [MICS6 Manual for Statistical Data Analysis](https://data.unicef.org/wp-content/uploads/2019/07/MICS6-manual-for-stats-data-analysis-English_2020_v2.pdf).

The conceptual keys are in [Files, units of observation and identifiers](#c02-files). The three common links are:

1. Household information onto the selected child: `HH1 + HH2`.
2. The selected child's own household-list record: `fs` uses `HH1 + HH2 + LN`; `hl` uses `HH1 + HH2 + HL1`.
3. Another household member, such as the mother or caretaker recorded in `FS4`: `HH1 + HH2` plus that person's line number, not the child's own `LN`.

### Example 1: add household variables to `fs`

Suppose we want to add household variables from `hh.sav`.

#### R

```{r eval=FALSE}
library(haven)
library(dplyr)

fs <- read_sav("Data/UNICEF/COUNTRY_SURVEY/fs.sav")
hh <- read_sav("Data/UNICEF/COUNTRY_SURVEY/hh.sav")

hh %>%
  count(HH1, HH2) %>%
  filter(n > 1)

hh_keep <- hh %>%
  select(HH1, HH2, variable_1, variable_2)

fs_merged <- fs %>%
  left_join(
    hh_keep,
    by = c("HH1", "HH2"),
    relationship = "many-to-one"
  )
```

Replace `variable_1` and `variable_2` with the variables required for the analysis.

`relationship = "many-to-one"` makes the expectation explicit: many child records may refer to a household, but there should be only one matching household record for each `HH1 + HH2` key.

#### Stata

```stata
use "Data/UNICEF/COUNTRY_SURVEY/fs.dta", clear

merge m:1 HH1 HH2 using ///
    "Data/UNICEF/COUNTRY_SURVEY/hh.dta", ///
    keepusing(variable_1 variable_2)

tab _merge
```

If you are working directly with UNICEF's `.sav` files in Stata, import the SPSS files first and save temporary `.dta` versions.

### Example 2: add the child's own `hl` variables

The line-number variable has a different name in the two files:

```text
fs: LN
hl: HL1
```

#### R

```{r eval=FALSE}
library(haven)
library(dplyr)

fs <- read_sav("Data/UNICEF/COUNTRY_SURVEY/fs.sav")
hl <- read_sav("Data/UNICEF/COUNTRY_SURVEY/hl.sav")

hl %>%
  count(HH1, HH2, HL1) %>%
  filter(n > 1)

hl_child <- hl %>%
  select(HH1, HH2, HL1, variable_1, variable_2)

fs_merged <- fs %>%
  left_join(
    hl_child,
    by = c("HH1", "HH2", "LN" = "HL1"),
    relationship = "one-to-one"
  )
```

Because there is only one selected child per household in `fs`, the child-level link should be one-to-one.

#### Stata

Rename the line number in the using file:

```stata
import spss using ///
    "Data/UNICEF/COUNTRY_SURVEY/hl.sav", clear

keep HH1 HH2 HL1 variable_1 variable_2
rename HL1 LN

isid HH1 HH2 LN

tempfile hl_child
save `hl_child'

import spss using ///
    "Data/UNICEF/COUNTRY_SURVEY/fs.sav", clear

isid HH1 HH2 LN

merge 1:1 HH1 HH2 LN using `hl_child'

tab _merge
```

### Always validate a merge

A merge command that completes without an error is not evidence that the merge is correct.

Before merging, ask:

1. What does one row represent in each file?
2. Should the key be unique in one file or both?
3. Are the line numbers identifying the same person?
4. How many rows should the analysis dataset have after the merge?

Then inspect the result.

#### R checks

```{r eval=FALSE}
nrow(fs)
nrow(fs_merged)

fs_merged %>%
  count(HH1, HH2, LN) %>%
  filter(n > 1)
```

For important merges, use an indicator to identify unmatched observations rather than silently accepting them.

```{r eval=FALSE}
fs_check <- fs %>%
  left_join(
    hl %>%
      select(HH1, HH2, HL1, variable_1),
    by = c("HH1", "HH2", "LN" = "HL1"),
    relationship = "one-to-one",
    unmatched = "drop"
  )
```

For production work, count unmatched records rather than simply dropping them.

#### Stata checks

Stata creates `_merge` automatically:

```stata
tab _merge
```

Interpret:

```text
1 = record only in master
2 = record only in using file
3 = matched
```

Do not automatically write:

```stata
drop if _merge != 3
```

until you understand why observations did not match.

### A merge can change neither the unit of analysis nor the weight

If the dataset still contains one row per selected `fs` child after adding household variables, the analysis population remains the selected-child population.

```text
fs outcome + hh wealth variable
          │
          ▼
one row per selected child
          │
          ▼
use fsweight
```

You do not switch to `hhweight` because one covariate came from the household file. See [Survey design, weights and standard errors](#c05-survey-design).

### AFLEARN scripts already perform some required links

You do not necessarily need to perform every merge manually.

AFLEARN's cross-country `fs` harmonisation script reads `hl.sav` where needed to add selected-child household-list variables such as father education and school type. The script checks `HH1`, `HH2` and line-number identifiers as part of that workflow.

The examples above are included so that you understand the structure and can add variables that are not already covered by AFLEARN's harmonised file.
