# Merging MICS6 files within a country {#c03-merge}

MICS data are split across files because the files have different units of observation. You will often need to add information from another file to the selected-child (`fs`) data.

The most important rule is:

::: {.alternativebox-solo}
**Decide what one row of your final dataset represents before you merge anything.**
:::

For foundational-learning research, one row will usually represent the **selected child in `fs`**. The safest workflow is therefore generally to start with `fs` and add the variables you need.

UNICEF's MICS6 statistical analysis manual includes specific guidance on merging MICS datasets and describes the identifiers used across questionnaire files. See the [MICS6 Manual for Statistical Data Analysis](https://data.unicef.org/wp-content/uploads/2019/07/MICS6-manual-for-stats-data-analysis-English_2020_v2.pdf).

## Three common types of link

### 1. Add household information to the selected child

Use:

```text
HH1 + HH2
```

Conceptually:

```text
fs child                         hh household
HH1  HH2  LN                     HH1  HH2
101   01   3  ───────────────►   101   01
```

This is appropriate for variables that genuinely have **one value per household**.

### 2. Add the selected child's own household-list information

Use:

```text
fs: HH1 + HH2 + LN
hl: HH1 + HH2 + HL1
```

where the selected child's `LN` corresponds to their `HL1` in the household listing.

### 3. Link the child to another household member

Some questionnaire variables store another person's household line number. For example, `FS4` identifies the child's mother/caretaker in the standard MICS6 questionnaire.

In that case the key is:

```text
HH1 + HH2 + other person's line number
```

not the child's own `LN`.

See [Files, units of observation and identifiers](#c02-files) for the conceptual explanation.

# Example 1: add household variables to `fs`

Suppose we want to add household variables from `hh.sav`.

## R

```{r eval=FALSE}
library(haven)
library(dplyr)

fs <- read_sav("Data/UNICEF/COUNTRY_SURVEY/fs.sav")
hh <- read_sav("Data/UNICEF/COUNTRY_SURVEY/hh.sav")

# Check that there is only one household record per key
hh %>%
  count(HH1, HH2) %>%
  filter(n > 1)

# Keep only the household variables needed
hh_keep <- hh %>%
  select(HH1, HH2, variable_1, variable_2)

# Add them to the selected-child records
fs_merged <- fs %>%
  left_join(
    hh_keep,
    by = c("HH1", "HH2"),
    relationship = "many-to-one"
  )
```

Replace `variable_1` and `variable_2` with the variables required for your analysis.

Using `relationship = "many-to-one"` makes your expectation explicit: many child/individual records may refer to a household, but there should be only one matching household record for each `HH1 + HH2` key.

## Stata

```stata
use "Data/UNICEF/COUNTRY_SURVEY/fs.dta", clear

merge m:1 HH1 HH2 using ///
    "Data/UNICEF/COUNTRY_SURVEY/hh.dta", ///
    keepusing(variable_1 variable_2)

tab _merge
```

If you are working directly with UNICEF's `.sav` files in Stata, import the SPSS files first and save temporary `.dta` versions.

# Example 2: add the child's own `hl` variables

The line-number variable has a different name in the two files:

```text
fs: LN
hl: HL1
```

## R

```{r eval=FALSE}
library(haven)
library(dplyr)

fs <- read_sav("Data/UNICEF/COUNTRY_SURVEY/fs.sav")
hl <- read_sav("Data/UNICEF/COUNTRY_SURVEY/hl.sav")

# Verify household-member keys are unique
hl %>%
  count(HH1, HH2, HL1) %>%
  filter(n > 1)

hl_child <- hl %>%
  select(HH1, HH2, HL1, variable_1, variable_2)

fs_merged <- fs %>%
  left_join(
    hl_child,
    by = c("HH1", "HH2", "LN" = "HL1"),
    relationship = "one-to-one"
  )
```

Because there is only one selected child per household in `fs`, the child-level link should be one-to-one.

## Stata

A simple approach is to rename the line number in the using file:

```stata
import spss using ///
    "Data/UNICEF/COUNTRY_SURVEY/hl.sav", clear

keep HH1 HH2 HL1 variable_1 variable_2
rename HL1 LN

isid HH1 HH2 LN

tempfile hl_child
save `hl_child'

import spss using ///
    "Data/UNICEF/COUNTRY_SURVEY/fs.sav", clear

isid HH1 HH2 LN

merge 1:1 HH1 HH2 LN using `hl_child'

tab _merge
```

# Always validate a merge

Do not regard a merge command completing without an error as evidence that the merge is correct.

Before merging, ask:

1. **What does one row represent in each file?**
2. **Should the key be unique in one file or both?**
3. **Are the line numbers identifying the same person?**
4. **How many rows should the analysis dataset have after the merge?**

Then inspect the result.

## R checks

```{r eval=FALSE}
# Number of rows should normally remain the same as fs
nrow(fs)
nrow(fs_merged)

# Child identifier should remain unique
fs_merged %>%
  count(HH1, HH2, LN) %>%
  filter(n > 1)
```

For important merges, use an indicator to identify unmatched observations rather than silently accepting them.

```{r eval=FALSE}
fs_check <- fs %>%
  left_join(
    hl %>%
      select(HH1, HH2, HL1, variable_1),
    by = c("HH1", "HH2", "LN" = "HL1"),
    relationship = "one-to-one",
    unmatched = "drop"
  )
```

For production work, we recommend explicitly counting unmatched records rather than simply dropping them.

## Stata checks

Stata creates `_merge` automatically:

```stata
tab _merge
```

Interpret:

```text
1 = record only in master
2 = record only in using file
3 = matched
```

Do not automatically write:

```stata
drop if _merge != 3
```

until you understand why observations did not match.

# A merge can change neither the unit of analysis nor the weight

If your dataset still contains one row per selected `fs` child after adding household variables, the analysis population remains the selected-child population.

For example:

```text
fs outcome + hh wealth variable
          │
          ▼
one row per selected child
          │
          ▼
use fsweight
```

You do **not** switch to `hhweight` because one covariate came from the household file.

See [Survey design, weights and standard errors](#c05-survey-design).

# AFLEARN scripts already perform some required links

You do not necessarily need to perform every merge manually.

For example, AFLEARN's cross-country `fs` harmonisation script reads `hl.sav` where needed to add selected-child household-list variables such as father education and school type. The script checks `HH1`, `HH2` and line-number identifiers as part of that workflow.

The manual examples above are included so that you understand the structure and can add variables that are not already covered by AFLEARN's harmonised file.

## Next

For multi-country analysis, continue to [Harmonising variables across countries](#c03-harmonise).
