## Harmonising variables across countries {#c03-harmonise}

The MICS6 questionnaires are standardised, but the released country datasets are not identical.

Differences can include:

- country-specific variable names;
- questionnaire adaptations;
- education-system codes;
- language codes;
- variables omitted because a module was not administered;
- country-specific response categories; and
- occasional questionnaire or data-layout anomalies.

Appending `fs.sav` files from several countries will not produce a reliable multi-country analysis file.

AFLEARN provides an R script that does the repetitive harmonisation work for the MICS6 surveys covered by this guide.

### What the AFLEARN script produces

The current script is [`harmonize-mics-fs-v1.0.R`](https://github.com/datafirst-aflearn/MICS6-Guide/blob/main/Script/harmonize-mics-fs-v1.0.R). It reads the survey folders in `Data/UNICEF/` and writes to `Data/AFLEARN Harmonised Data/`:

| Output | Purpose |
|---|---|
| `mics6-fs-harmonised.dta` and `.rds` | Harmonised child-level analysis file, readable in both R and Stata |
| `mics6_fs_variable_crosswalk.xlsx` | Source variables, harmonisation maps, exclusions and notes |

The existing AFLEARN implementation covers 18 African MICS6 country codes:

```text
BEN CAF COD COM GHA GMB GNB LSO MDG MWI NGA SLE STP SWZ TCD TGO TUN ZWE
```

Tunisia has more than one MICS6 survey year in the current processing system.

### What is harmonised?

The script creates a child-level file containing commonly used variables from the MICS6 5–17 questionnaire and selected linked information.

| Area | Examples |
|---|---|
| Identifiers | country, year, cluster, household, child line number |
| Survey design | `fsweight`, `fshweight` |
| Geography | urban/rural, region |
| Household | number of children aged 5–17 |
| Parents | mother and father education |
| Schooling | attendance, education level and grade |
| Child labour | work, chores and hours |
| Parental involvement | books and school engagement |
| Child functioning | functioning items |
| Discipline | discipline methods and attitudes |
| Foundational learning | numeracy items and FLS setup variables |

Reading outcomes are kept separate because they require additional passage-specific processing. They are constructed on the next page and can be merged onto this file using the child identifiers.

The current script retains `fsweight` but not `PSU`, `stratum` or `windex5`. Add those from the source files before survey-weighted analysis; see [Prepare the analysis data](#c06-prepare-analysis).

### How to re-run it

`run-it.R` already sources this script after unpacking the UNICEF download. To re-run only the FS harmonisation, from the project directory:

```{r eval=FALSE}
source("Script/harmonize-mics-fs-v1.0.R")
```

The project should look like:

```text
MICS project/
├── Data/
│   ├── UNICEF/
│   │   ├── BEN_2021_MICS6_v01_M/
│   │   │   ├── fs.sav
│   │   │   ├── hl.sav
│   │   │   └── ...
│   │   └── ...
│   └── AFLEARN Harmonised Data/
└── Script/
    └── harmonize-mics-fs-v1.0.R
```

The script uses R packages including `tidyverse`, `haven`, `labelled` and `openxlsx`.

The country-specific renaming rules, questionnaire anomalies and coding decisions are documented in [Technical documentation](#c07-technical), together with the variable crosswalk.

### A harmonised name does not always mean harmonised meaning

Some variables have a common name but intentionally retain country-specific codes.

Examples include:

- `region`;
- raw education levels and grades; and
- `school_type`.

Do not pool the numeric codes for these variables across countries without checking the documentation.

For education, AFLEARN also creates common level variables such as:

```text
current_level_h
highest_level_h
previous_level_h
mother_edu_h
father_edu_h
```

These provide a common broad education-level classification.

### Grade needs particular care

`current_grade` is the year within an education level, not the child's overall year of schooling.

```text
Primary, year 1       current_grade = 1
Secondary, year 1     current_grade = 1
```

Those children are not at the same point in the school system.

Use `current_grade` together with the harmonised education-level variable, or construct a system-level grade measure appropriate to the analysis. See [Why current_grade is not the child's grade](#c07-current-grade).

### Check the output

In R:

```{r eval=FALSE}
library(tidyverse)

d <- haven::read_dta("Data/AFLEARN Harmonised Data/mics6-fs-harmonised.dta")

d %>%
  count(country_iso3, year)

d %>%
  count(country_iso3, year, HH1, HH2, LN) %>%
  filter(n > 1)

d %>%
  count(country_iso3, current_level_h)

d %>%
  count(country_iso3, mother_edu_h)
```

In Stata:

```stata
use "Data/AFLEARN Harmonised Data/mics6-fs-harmonised.dta", clear

tab country_iso3
tab current_level_h country_iso3
tab mother_edu_h country_iso3

isid country_iso3 year HH1 HH2 LN
```

### Use the crosswalk

The Excel crosswalk is part of the research output. It is published with the harmonisation code in the [AFLEARN MICS6-Guide repository](https://github.com/datafirst-aflearn/MICS6-Guide). How to read it is in [Variable crosswalks and dictionaries](#c07-crosswalks).

If a variable matters substantively to the research question, check its crosswalk entry before pooling it across countries.

The harmonised `fs` file contains numeracy items and other FLS setup variables, but does not duplicate the research-ready reading outcomes. Continue to [Constructing research-ready reading variables](#c03-reading).
