# Adding the correct reading data to an IPUMS extract {#c04-ipums-reading}

This page shows how to combine an IPUMS MICS child extract with AFLEARN's research-ready MICS6 reading data.

The purpose is to retain the main advantage of IPUMS — harmonised multi-country background variables — without using incomplete reading data in countries that administered multiple passages.

In a multi-passage survey, children assessed in the passage IPUMS does not retain are absent from the harmonised reading variables. Because that absence is generated by the language in which a child was assessed, dropping observations with missing IPUMS reading variables does not solve the problem — it creates a systematically selected analysis sample. The limitation is set out in [Getting data from IPUMS MICS](#c01-ipums).

## AFLEARN's solution

AFLEARN constructs the reading variables from the original UNICEF MICS6 country files and retains all relevant passages.

The reading file includes variables such as:

```text
passage_language
passage_length
words_att
words_incorrect
reading_score
reading_accuracy
read_comp_1 ... read_comp_5
read_comp_score
reading_skills
reading_status
```

and passage-specific / multiple-attempt information where relevant.

See [Constructing research-ready reading variables](#c03-reading) for definitions.

## The linking identifiers

AFLEARN's reading file deliberately retains both the original UNICEF identifiers and IPUMS-style linking identifiers.

For linking to IPUMS, the relevant fields are:

```text
AFLEARN             IPUMS
--------------------------------
cluster        <->  CLUSTER
hhno           <->  HHNO
linech         <->  LINECH
survey/sample  <->  SAMPLE
```

IPUMS's official linking guidance confirms that `SAMPLE`, `CLUSTER`, `HHNO` and the person line number are the key identifying variables in pooled extracts.

The original MICS6 identifiers underlying these are:

```text
HH1 -> CLUSTER
HH2 -> HHNO
selected child's line number -> LINECH
```

## Step 1: create your IPUMS child extract

Choose:

```text
Unit of analysis: Children age 5–17
```

Select the MICS6 samples and background variables needed for your research.

Make sure the resulting extract retains:

```text
SAMPLE
CLUSTER
HHNO
LINECH
```

or their corresponding automatically included technical identifiers.

Run the IPUMS-provided Stata syntax to create the harmonised `.dta` file.

See the [IPUMS MICS extract instructions](https://mics.ipums.org/mics/extract_instructions.shtml).

## Step 2: obtain the AFLEARN reading file

Use the AFLEARN-provided research-ready reading file, or construct it from the original UNICEF data using:

```text
harmonize-mics-reading-v1.1.R
```

as described in [Constructing research-ready reading variables](#c03-reading).

## Step 3: make the sample identifier compatible

`CLUSTER`, `HHNO` and `LINECH` identify a child **within a survey**, but they are not sufficient for a pooled multi-country file because the same numbers occur in different surveys.

You must therefore include the survey/sample identifier in the merge.

AFLEARN's reading file contains country and year identifiers from which the corresponding IPUMS sample can be mapped.

We recommend creating an explicit crosswalk:

```text
AFLEARN country + year     IPUMS SAMPLE
----------------------------------------
Malawi 2019/20             [sample code]
Benin 2021                 [sample code]
...
```

and storing that crosswalk with the project.

::: {.warningbox-header}
**Do not guess IPUMS sample codes from country names**
:::

::: {.warningbox}
Use the actual `SAMPLE` or `SAMPLE_STR` values in your extract and validate the mapping once. IPUMS sample identifiers distinguish survey year and, where necessary, national and subnational samples.
:::

## Step 4: check uniqueness before merging

### R

```{r eval=FALSE}
library(haven)
library(dplyr)

ipums <- read_dta("Data/IPUMS/ipums_mics_fs.dta")
reading <- read_dta("Data/AFLEARN Harmonised Data/mics6_reading_harmonized.dta")

ipums %>%
  count(SAMPLE, CLUSTER, HHNO, LINECH) %>%
  filter(n > 1)
```

The AFLEARN file should be checked using the corresponding mapped sample identifier:

```{r eval=FALSE}
reading %>%
  count(SAMPLE, cluster, hhno, linech) %>%
  filter(n > 1)
```

If either check returns observations, investigate before merging.

### Stata

```stata
use "Data/IPUMS/ipums_mics_fs.dta", clear

duplicates report SAMPLE CLUSTER HHNO LINECH
```

and perform the corresponding check on the AFLEARN file after adding the IPUMS sample identifier.

## Step 5: merge in R

Assuming the AFLEARN file has been given the matching `SAMPLE` value:

```{r eval=FALSE}
library(haven)
library(dplyr)

ipums <- read_dta("Data/IPUMS/ipums_mics_fs.dta")

reading <- read_dta(
  "Data/AFLEARN Harmonised Data/mics6_reading_harmonized.dta"
)

analysis <- ipums %>%
  left_join(
    reading,
    by = c(
      "SAMPLE",
      "CLUSTER" = "cluster",
      "HHNO" = "hhno",
      "LINECH" = "linech"
    ),
    relationship = "one-to-one"
  )
```

The `left_join()` keeps the IPUMS child extract as the analysis population and adds AFLEARN reading information where available.

## Step 6: merge in Stata

Rename the AFLEARN identifiers to match IPUMS:

```stata
use "Data/AFLEARN Harmonised Data/mics6_reading_harmonized.dta", clear

rename cluster CLUSTER
rename hhno HHNO
rename linech LINECH

* SAMPLE must already have been mapped to the
* corresponding IPUMS sample identifier.

isid SAMPLE CLUSTER HHNO LINECH

tempfile reading
save `reading'

use "Data/IPUMS/ipums_mics_fs.dta", clear

merge 1:1 SAMPLE CLUSTER HHNO LINECH using `reading'

tab _merge
```

IPUMS's own guidance uses the same underlying approach for linking additional variables from original UNICEF MICS files to an IPUMS extract.

See [IPUMS MICS Linking](https://mics.ipums.org/mics/linking.shtml).

## Step 7: inspect the merge

Do not immediately drop unmatched records.

In Stata:

```stata
tab _merge SAMPLE
```

In R, create an explicit match flag:

```{r eval=FALSE}
analysis <- ipums %>%
  left_join(
    reading %>%
      mutate(aflearn_reading_match = TRUE),
    by = c(
      "SAMPLE",
      "CLUSTER" = "cluster",
      "HHNO" = "hhno",
      "LINECH" = "linech"
    ),
    relationship = "one-to-one"
  )

analysis %>%
  count(SAMPLE, aflearn_reading_match)
```

Remember that not every `fs` child is expected to have a completed reading assessment. Children aged 5–6 and 15–17 are outside the FLS target age range, and eligible children can also follow other reading-status pathways.

The important question is whether the **child record links correctly**, not whether every child has a non-missing reading score.

## Step 8: remove or clearly distinguish the IPUMS reading variables

For multi-passage surveys, we recommend removing the incomplete IPUMS reading outcome variables from the final analysis dataset once the AFLEARN variables have been added.

This prevents accidental use later.

For example, in R:

```{r eval=FALSE}
analysis <- analysis %>%
  select(
    -FLWORDSATTEMPT
    # add other IPUMS reading variables to remove here
  )
```

or rename them clearly if you need to retain them for validation.

In Stata:

```stata
drop FLWORDSATTEMPT
```

The final production script should list the complete set of IPUMS reading variables that should be excluded or replaced.

::: {.alternativebox-header}
**Recommended principle**
:::

::: {.alternativebox}
In multi-passage countries, there should be only one clearly identified set of reading outcomes in the analysis file: the AFLEARN variables constructed from the complete UNICEF source data.
:::

## Step 9: validate the combined file

At minimum, check:

### Number of children by survey

```{r eval=FALSE}
analysis %>%
  count(SAMPLE)
```

### Reading status among children aged 7–14

```{r eval=FALSE}
analysis %>%
  filter(age >= 7, age <= 14) %>%
  count(SAMPLE, reading_status)
```

Use the actual IPUMS age variable selected in your extract.

### Passage language

```{r eval=FALSE}
analysis %>%
  count(SAMPLE, passage_language)
```

For a known multi-passage country such as Malawi, this is an important validation: both relevant assessment languages should be represented in the AFLEARN reading data.

### Compare against the original UNICEF processing

For supported surveys, aggregate results from the merged IPUMS + AFLEARN file should reproduce the corresponding results from the AFLEARN/UNICEF child-level analysis file, allowing for any intentional differences in the analysis sample.

## A useful validation exercise for Malawi

The Malawi case is a particularly clear test of whether the merge has worked.

Before replacement, the IPUMS reading data identify only the retained English-passage observations in `FLWORDSATTEMPT`.

After adding the AFLEARN reading file, the analysis should recover the Chichewa assessments as well.

A simple validation table should show, by passage language:

```text
Malawi
-----------------------------
English       ...
Chichewa      ...
Both          ...
```

and should reconcile with the original UNICEF source data.

## What if I use only single-passage countries?

If your analysis is restricted entirely to surveys with one reading passage, the specific multi-passage omission described here does not arise.

However, AFLEARN's reading variables still offer advantages:

- a consistent reading-status variable;
- harmonised passage metadata;
- consistent accuracy construction; and
- the same reading definitions across all supported MICS6 surveys.

Using the AFLEARN reading file for all countries can therefore simplify a project and avoid maintaining different reading workflows for different surveys.

## Next

Once the IPUMS and AFLEARN data are combined, the next major issue is statistical rather than data management.

Continue to [Survey design, weights and standard errors](#c05-survey-design) before calculating population estimates, confidence intervals or regression standard errors.
