## Adding the correct reading data to an IPUMS extract {#c04-ipums-reading}

This page shows how to combine an IPUMS MICS child extract with AFLEARN's research-ready MICS6 reading data.

The purpose is to retain the main advantage of IPUMS — harmonised multi-country background variables — without using incomplete reading data in countries that administered multiple passages.

In a multi-passage survey, children assessed in the passage IPUMS does not retain are absent from the harmonised reading variables. Because that absence is generated by the language in which a child was assessed, dropping observations with missing IPUMS reading variables does not solve the problem: it creates a systematically selected analysis sample.

### AFLEARN's solution

AFLEARN constructs the reading variables from the original UNICEF MICS6 country files and retains all relevant passages.

The reading file includes:

```text
passage_language
passage_length
words_att
words_incorrect
reading_score
reading_accuracy
read_comp_1 ... read_comp_5
read_comp_score
reading_skills
reading_status
```

and passage-specific / multiple-attempt information where relevant.

See [Constructing research-ready reading variables](#c03-reading) for definitions.

### Step 1: create your IPUMS child extract

Choose:

```text
Unit of analysis: Children age 5–17
```

Select the MICS6 samples and background variables needed for the research.

Make sure the resulting extract retains:

```text
SAMPLE
CLUSTER
HHNO
LINECH
```

or their corresponding automatically included technical identifiers.

Run the IPUMS-provided Stata syntax to create the harmonised `.dta` file.

See the [IPUMS MICS extract instructions](https://mics.ipums.org/mics/extract_instructions.shtml).

### Step 2: obtain the AFLEARN reading file

Use the AFLEARN-provided research-ready reading file, or construct it from the original UNICEF data using `harmonize-mics-reading-v1.1.R` as described in [Constructing research-ready reading variables](#c03-reading).

### Step 3: make the sample identifier compatible

`CLUSTER`, `HHNO` and `LINECH` identify a child within a survey, but they are not sufficient for a pooled multi-country file because the same numbers occur in different surveys.

Include the survey/sample identifier in the merge.

AFLEARN's reading file contains country and year identifiers from which the corresponding IPUMS sample can be mapped.

Create an explicit crosswalk:

```text
AFLEARN country + year     IPUMS SAMPLE
----------------------------------------
Malawi 2019/20             [sample code]
Benin 2021                 [sample code]
...
```

and store that crosswalk with the project.

::: {.warningbox-header}
**Do not guess IPUMS sample codes from country names**
:::

::: {.warningbox}
Use the actual `SAMPLE` or `SAMPLE_STR` values in your extract and validate the mapping once. IPUMS sample identifiers distinguish survey year and, where necessary, national and subnational samples.
:::

### Step 4: check uniqueness before merging

#### R

```{r eval=FALSE}
library(haven)
library(dplyr)

ipums <- read_dta("Data/ipums_mics_fs.dta")
reading <- read_dta("Data/AFLEARN Harmonised Data/mics6-reading-harmonised.dta")

ipums %>%
  count(SAMPLE, CLUSTER, HHNO, LINECH) %>%
  filter(n > 1)
```

The AFLEARN file should be checked using the corresponding mapped sample identifier:

```{r eval=FALSE}
reading %>%
  count(SAMPLE, cluster, hhno, linech) %>%
  filter(n > 1)
```

If either check returns observations, investigate before merging.

#### Stata

```stata
use "Data/ipums_mics_fs.dta", clear

duplicates report SAMPLE CLUSTER HHNO LINECH
```

and perform the corresponding check on the AFLEARN file after adding the IPUMS sample identifier.

### Step 5: merge in R

Assuming the AFLEARN file has been given the matching `SAMPLE` value:

```{r eval=FALSE}
library(haven)
library(dplyr)

ipums <- read_dta("Data/ipums_mics_fs.dta")

reading <- read_dta(
  "Data/AFLEARN Harmonised Data/mics6-reading-harmonised.dta"
)

analysis <- ipums %>%
  left_join(
    reading,
    by = c(
      "SAMPLE",
      "CLUSTER" = "cluster",
      "HHNO" = "hhno",
      "LINECH" = "linech"
    ),
    relationship = "one-to-one"
  )
```

The `left_join()` keeps the IPUMS child extract as the analysis population and adds AFLEARN reading information where available.

### Step 6: merge in Stata

Rename the AFLEARN identifiers to match IPUMS:

```stata
use "Data/AFLEARN Harmonised Data/mics6-reading-harmonised.dta", clear

rename cluster CLUSTER
rename hhno HHNO
rename linech LINECH

* SAMPLE must already have been mapped to the
* corresponding IPUMS sample identifier.

isid SAMPLE CLUSTER HHNO LINECH

tempfile reading
save `reading'

use "Data/ipums_mics_fs.dta", clear

merge 1:1 SAMPLE CLUSTER HHNO LINECH using `reading'

tab _merge
```

IPUMS's own guidance uses the same underlying approach for linking additional variables from original UNICEF MICS files to an IPUMS extract.

See [IPUMS MICS Linking](https://mics.ipums.org/mics/linking.shtml).

### Step 7: inspect the merge

Do not immediately drop unmatched records.

In Stata:

```stata
tab _merge SAMPLE
```

In R, create an explicit match flag:

```{r eval=FALSE}
analysis <- ipums %>%
  left_join(
    reading %>%
      mutate(aflearn_reading_match = TRUE),
    by = c(
      "SAMPLE",
      "CLUSTER" = "cluster",
      "HHNO" = "hhno",
      "LINECH" = "linech"
    ),
    relationship = "one-to-one"
  )

analysis %>%
  count(SAMPLE, aflearn_reading_match)
```

Not every `fs` child is expected to have a completed reading assessment. Children aged 5–6 and 15–17 are outside the FLS target age range, and eligible children can also follow other reading-status pathways.

The important question is whether the child record links correctly, not whether every child has a non-missing reading score.

### Step 8: remove or clearly distinguish the IPUMS reading variables

For multi-passage surveys, remove the incomplete IPUMS reading outcome variables from the final analysis dataset once the AFLEARN variables have been added.

This prevents accidental use later.

For example, in R:

```{r eval=FALSE}
analysis <- analysis %>%
  select(
    -FLWORDSATTEMPT
    # add other IPUMS reading variables to remove here
  )
```

or rename them clearly if you need to retain them for validation.

In Stata:

```stata
drop FLWORDSATTEMPT
```

The final production script should list the complete set of IPUMS reading variables that should be excluded or replaced.

::: {.alternativebox-header}
**Recommended principle**
:::

::: {.alternativebox}
In multi-passage countries, there should be only one clearly identified set of reading outcomes in the analysis file: the AFLEARN variables constructed from the complete UNICEF source data.
:::
