## Validating the combined file {#c04-validate}

At minimum, check the following after merging AFLEARN reading onto an IPUMS child extract.

### Number of children by survey

```{r eval=FALSE}
analysis %>%
  count(SAMPLE)
```

### Reading status among children aged 7–14

```{r eval=FALSE}
analysis %>%
  filter(age >= 7, age <= 14) %>%
  count(SAMPLE, reading_status)
```

Use the actual IPUMS age variable selected in your extract.

### Passage language

```{r eval=FALSE}
analysis %>%
  count(SAMPLE, passage_language)
```

For a known multi-passage country such as Malawi, both relevant assessment languages should be represented in the AFLEARN reading data.

### Compare against the original UNICEF processing

For supported surveys, aggregate results from the merged IPUMS + AFLEARN file should reproduce the corresponding results from the AFLEARN/UNICEF child-level analysis file, allowing for any intentional differences in the analysis sample.

### A useful validation exercise for Malawi

The Malawi case is a particularly clear test of whether the merge has worked.

Before replacement, the IPUMS reading data identify only the retained English-passage observations in `FLWORDSATTEMPT`.

After adding the AFLEARN reading file, the analysis should recover the Chichewa assessments as well.

A simple validation table should show, by passage language:

```text
Malawi
-----------------------------
English       ...
Chichewa      ...
Both          ...
```

and should reconcile with the original UNICEF source data. The source counts are in [Reading-variable construction](#c07-reading-construction).

### What if I use only single-passage countries?

If the analysis is restricted entirely to surveys with one reading passage, the specific multi-passage omission described here does not arise.

AFLEARN's reading variables still offer:

- a consistent reading-status variable;
- harmonised passage metadata;
- consistent accuracy construction; and
- the same reading definitions across all supported MICS6 surveys.

Using the AFLEARN reading file for all countries can therefore simplify a project and avoid maintaining different reading workflows for different surveys.

Once the IPUMS and AFLEARN data are combined, the next issue is statistical rather than data management. Continue to [Survey design, weights and standard errors](#c05-survey-design) before calculating population estimates, confidence intervals or regression standard errors.
