# Setting up one MICS6 survey {#c05-single-survey}

For a single MICS6 survey, the standard selected-child design uses:

```text
weight   = fsweight
PSU      = PSU
stratum  = stratum
```

This page shows how to set that design up in R and Stata.

## First check the design variables

Before creating a survey object:

```{r eval=FALSE}
library(haven)
library(dplyr)

d <- read_dta("Data/AFLEARN Harmonised Data/mics6_fs_harmonized.dta")

d %>%
  summarise(
    n = n(),
    missing_weight = sum(is.na(fsweight)),
    missing_psu = sum(is.na(PSU)),
    missing_stratum = sum(is.na(stratum)),
    n_psu = n_distinct(PSU),
    n_strata = n_distinct(stratum)
  )
```

If you are using the AFLEARN harmonised file, confirm the final exact capitalisation of `PSU` in the production version.

You should also check how many PSUs occur within each stratum:

```{r eval=FALSE}
d %>%
  distinct(stratum, PSU) %>%
  count(stratum, name = "n_psu") %>%
  arrange(n_psu)
```

## R: the `survey` package

```{r eval=FALSE}
library(survey)

des <- svydesign(
  ids = ~PSU,
  strata = ~stratum,
  weights = ~fsweight,
  data = d,
  nest = TRUE
)
```

`nest = TRUE` tells `survey` that PSU identifiers are nested within strata. This is a safe specification when PSU codes can repeat between strata.

A weighted mean:

```{r eval=FALSE}
svymean(
  ~pct_total_math,
  design = des,
  na.rm = TRUE
)
```

A proportion for a 0/1 indicator:

```{r eval=FALSE}
svymean(
  ~foundational_numeracy,
  design = des,
  na.rm = TRUE
)
```

## R: `srvyr` for tidyverse-style analysis

If you prefer `dplyr` syntax, `srvyr` wraps the `survey` package.

```{r eval=FALSE}
library(srvyr)

des <- d %>%
  as_survey_design(
    ids = PSU,
    strata = stratum,
    weights = fsweight,
    nest = TRUE
  )
```

Then:

```{r eval=FALSE}
des %>%
  summarise(
    mean_math = survey_mean(
      pct_total_math,
      na.rm = TRUE,
      vartype = c("se", "ci")
    )
  )
```

or by a subgroup:

```{r eval=FALSE}
des %>%
  group_by(sex) %>%
  summarise(
    mean_math = survey_mean(
      pct_total_math,
      na.rm = TRUE,
      vartype = c("se", "ci")
    )
  )
```

## Stata

```stata
svyset PSU [pw=fsweight], strata(stratum)
```

Check the design:

```stata
svydescribe
```

Then estimate, for example:

```stata
svy: mean pct_total_math
```

For a 0/1 foundational-numeracy indicator:

```stata
svy: mean foundational_numeracy
```

The mean of a variable coded 0/1 is the estimated proportion equal to 1.

## Why not use `HH1` automatically as the PSU?

MICS datasets contain both the original cluster identifier (`HH1`) and design variables such as `PSU`.

For many surveys these may correspond closely, but the safest approach is to use the **sampling-error design variables supplied with the released data**.

In the standard MICS6 naming convention:

```text
PSU      primary sampling unit
stratum  sample stratum
```

IPUMS also harmonises these as `PSU` and `STRATUM`.

## Lonely PSUs

Variance estimation within a stratum relies on having more than one sampled PSU in that stratum.

Occasionally, after restricting to a domain or because of a particular survey design, software may encounter a stratum with only one contributing PSU.

In R, the `survey` package provides several options for handling these "lonely PSUs". Do not choose one mechanically. First determine whether the problem exists in the full survey design or only after defining a subpopulation.

A useful default for analysis workflows where lonely domains occur is:

```{r eval=FALSE}
options(
  survey.lonely.psu = "adjust",
  survey.adjust.domain.lonely = TRUE
)
```

but the chosen treatment should be documented.

In Stata, inspect singleton-unit messages and the survey design rather than silently ignoring them.

## Do not filter the raw data before defining the survey design

This is especially important when estimating a subgroup.

Suppose you want results for children aged 7–14.

Avoid:

```{r eval=FALSE}
# Not recommended for domain inference
d_7_14 <- d %>%
  filter(age >= 7, age <= 14)

des <- svydesign(
  ids = ~PSU,
  strata = ~stratum,
  weights = ~fsweight,
  data = d_7_14,
  nest = TRUE
)
```

Instead, define the survey design using the full relevant sample and then restrict the **design object**:

```{r eval=FALSE}
des <- svydesign(
  ids = ~PSU,
  strata = ~stratum,
  weights = ~fsweight,
  data = d,
  nest = TRUE
)

des_7_14 <- subset(
  des,
  age >= 7 & age <= 14
)
```

With `srvyr`:

```{r eval=FALSE}
des_7_14 <- des %>%
  filter(age >= 7, age <= 14)
```

This preserves the information survey software needs for correct domain/subpopulation variance estimation.

## Stata subpopulations

The same principle applies in Stata.

Rather than physically dropping everyone outside the domain and then estimating:

```stata
svy, subpop(if age >= 7 & age <= 14): ///
    mean pct_total_math
```

For a more complex domain, create a 0/1 indicator:

```stata
gen age7_14 = inrange(age,7,14)

svy, subpop(age7_14): mean pct_total_math
```

`subpop()` tells Stata that the target is a subpopulation of the original survey sample.

## Next

If your analysis contains more than one MICS survey, additional steps are required. Continue to [Combining MICS6 surveys](#c05-pooling).
