# Combining MICS6 surveys {#c05-pooling}

Multi-country MICS analysis introduces two separate questions:

1. **How should the survey design be represented technically?**
2. **What population does a pooled estimate represent?**

These should not be confused.

Making PSU identifiers unique across countries solves a **variance-estimation problem**. It does not, by itself, make a pooled weighted mean a meaningful regional estimate.

## Problem 1: PSU and stratum numbers repeat across surveys

Suppose two surveys both contain:

```text
PSU = 1
stratum = 1
```

They are obviously not the same sampling unit or stratum.

If surveys are appended without changing these identifiers, survey software can incorrectly treat unrelated observations as belonging to the same design units.

The solution is to create identifiers that are unique to each survey.

## Create a survey identifier

For AFLEARN's UNICEF-based harmonised data, a survey identifier can be based on:

```text
country_iso3 + year
```

For IPUMS data, use the IPUMS sample identifier such as `SAMPLE`.

Then create survey-specific PSU and stratum variables.

## R

```{r eval=FALSE}
library(dplyr)

d <- d %>%
  mutate(
    survey_id = interaction(
      country_iso3,
      year,
      drop = TRUE
    ),
    psu_pool = interaction(
      survey_id,
      PSU,
      drop = TRUE
    ),
    stratum_pool = interaction(
      survey_id,
      stratum,
      drop = TRUE
    )
  )
```

Then:

```{r eval=FALSE}
library(survey)

des_pool <- svydesign(
  ids = ~psu_pool,
  strata = ~stratum_pool,
  weights = ~fsweight,
  data = d,
  nest = TRUE
)
```

## Stata

```stata
egen survey_id = group(country_iso3 year)
egen psu_pool = group(survey_id PSU)
egen stratum_pool = group(survey_id stratum)

svyset psu_pool [pw=fsweight], strata(stratum_pool)
```

This tells the software correctly that a PSU in Malawi is not the same PSU as one carrying the same numeric code in Ghana.

::: {.alternativebox-header}
**Always make both PSU and stratum identifiers survey-specific in appended data**
:::

::: {.alternativebox}
Doing only the PSU is not enough if stratum codes also repeat across surveys.
:::

## Estimating each country separately

If the objective is:

> What is the mean numeracy score in each country?

then you are estimating **country-specific statistics**, not one pooled population parameter.

One transparent approach is to retain the combined design and estimate by country/survey.

### R with `srvyr`

```{r eval=FALSE}
des_pool %>%
  group_by(country_iso3) %>%
  summarise(
    mean_math = survey_mean(
      pct_total_math,
      na.rm = TRUE,
      vartype = c("se", "ci")
    )
  )
```

### Stata

```stata
svy: mean pct_total_math, over(country_iso3)
```

or run country-specific estimates while retaining the correct survey design.

For cross-country figures and tables, this is often exactly what you want.

## Problem 2: what does a pooled mean mean?

Suppose instead you calculate:

```text
one mean across all 18 countries
```

What population is that intended to describe?

There are several possibilities:

- the combined population of children in the 18 countries;
- the "average country", giving each country equal weight;
- the average child represented by the collection of normalised survey samples; or
- some externally defined regional population.

These are **different estimands** and require different weighting choices.

## MICS country weights do not automatically create a population-weighted regional estimate

MICS sample weights are constructed for inference within each survey and are generally normalised within the survey.

If you simply append countries and use the released `fsweight` values:

```text
pooled mean using fsweight
```

the relative contribution of one country versus another is driven partly by survey sample sizes and each survey's weight normalisation.

It should **not automatically be interpreted as the proportion of all children across those countries**.

For a true population-weighted regional estimate, you would need external population information appropriate to the target population and survey reference periods, and construct weights accordingly.

## Equal-country averages are another option

Sometimes the research question is:

> What does the average result across these countries look like if every country contributes equally?

A simple way is:

1. estimate the statistic separately within each country using the correct MICS design;
2. combine the country estimates with equal country weights.

That estimand is an **average of country estimates**, not an estimate for the pooled child population.

It should be labelled accordingly.

## Regression across countries

The same conceptual issue applies to pooled regressions.

Creating `psu_pool` and `stratum_pool` correctly handles the fact that design units belong to different surveys.

But you still need to decide:

- what the observation weights imply about the relative contribution of countries;
- whether country fixed effects are needed;
- whether coefficients are intended to describe within-country or between-country variation; and
- whether the target is an average relationship across the included surveys or a population-level regional relationship.

There is no single automatic "MICS pooled weight" that answers every multi-country research question.

::: {.alternativebox-header}
**Separate design from estimand**
:::

::: {.alternativebox}
`psu_pool` and `stratum_pool` tell the software **how the observations were sampled**.

Your treatment of country weights tells the analysis **what pooled population or average you are trying to represent**.
:::

## Repeated surveys from the same country

If you include two MICS6 surveys from the same country, such as different years or samples, the identifiers must still be **survey-specific**, not merely country-specific.

Use:

```text
country + year/sample
```

rather than country alone when creating pooled PSU and stratum identifiers.

## Recommended workflow for comparative AFLEARN analyses

For most descriptive comparisons in this guide:

1. retain the released `fsweight`;
2. make PSU and stratum identifiers unique by survey;
3. define the combined survey design;
4. estimate statistics **separately by country/survey**; and
5. plot or compare those country estimates.

Only create a single pooled regional statistic when the estimand and the required between-country weighting have been explicitly defined.
