# Getting standard errors right {#c05-standard-errors}

A point estimate and its uncertainty answer different questions.

For example:

```text
Estimated foundational numeracy = 23%
```

tells us the estimated population proportion.

A standard error tells us how precisely that proportion has been estimated under the sample design.

MICS analyses should therefore use design-based standard errors that account for:

- the selected-child weight;
- clustering within PSUs; and
- stratification.

## A quick comparison

Consider these three approaches:

```text
1. unweighted mean + ordinary SE
2. weighted mean + ordinary SE
3. survey-weighted mean + design-based SE
```

For population inference from MICS, the third is generally the appropriate approach.

The first ignores unequal selection probabilities.

The second incorporates weights into the point estimate but still ignores the clustered and stratified design in the variance calculation.

## Confidence intervals

A 95% confidence interval combines the estimate and its standard error.

In R with `srvyr`:

```{r eval=FALSE}
des %>%
  summarise(
    numeracy = survey_mean(
      foundational_numeracy,
      na.rm = TRUE,
      vartype = c("se", "ci")
    )
  )
```

In Stata:

```stata
svy: mean foundational_numeracy
```

Both return design-based uncertainty once the survey design has been specified correctly.

## Subgroups can become imprecise quickly

MICS surveys are designed to produce reliable estimates for specified populations and reporting domains. A dataset containing thousands of children nationally can have far fewer effective observations once you estimate, for example:

```text
Grade 2
× rural
× poorest quintile
× one country
```

Do not judge precision only from the raw number of child records.

Check:

- number of sampled PSUs contributing to the estimate;
- standard error;
- confidence interval;
- and, where useful, coefficient of variation or design effect.

## Design effects

The **design effect** compares the variance under the actual complex design with a reference variance under simple random sampling.

Values above 1 indicate that the complex design produces a larger variance than the simple-random-sample reference.

Clustering often increases variance because children within the same PSU are correlated. Stratification can improve precision.

In R, many `survey`/`srvyr` estimators can return a design effect.

For example:

```{r eval=FALSE}
svymean(
  ~pct_total_math,
  design = des,
  na.rm = TRUE,
  deff = TRUE
)
```

Design effects are useful diagnostics, but they are not a substitute for actually specifying the survey design.

## Degrees of freedom

For complex survey inference, precision depends strongly on the number of independent sampled PSUs, not simply the number of child observations.

Survey software therefore uses design-based degrees of freedom when constructing tests and confidence intervals.

This is another reason that a regression with several thousand child records may still have much less independent information than the raw `N` suggests.

## Missing outcomes

If an outcome is missing, survey estimators generally omit those observations for that particular estimate when `na.rm = TRUE` or the corresponding Stata command handles missing values.

But with reading, **why the outcome is missing matters substantively**.

Before estimating reading performance, use AFLEARN's `reading_status` to understand whether children:

- were outside the target age;
- did not have consent;
- faced an assessment-language mismatch;
- did not proceed through the reading sequence; or
- attempted the passage.

Do not treat all missing reading values as a generic statistical nuisance.

## Domain analysis, again

Suppose the target is:

> Mean numeracy among out-of-school children aged 10–14.

The safest sequence is:

```text
full selected-child survey data
          │
          ▼
define survey design
          │
          ▼
define age/schooling subpopulation
          │
          ▼
estimate mean + design-based SE
```

rather than:

```text
drop everyone else from raw data
          │
          ▼
pretend remaining observations
were the original survey sample
```

In R, subset/filter the survey **design object**.

In Stata, use `subpop()`.

## Regressions also need the survey design

The issue does not disappear when you move from descriptive statistics to regression.

In R:

```{r eval=FALSE}
model <- svyglm(
  pct_total_math ~ age + sex + wealth_quintile,
  design = des
)

summary(model)
```

In Stata:

```stata
svy: regress pct_total_math ///
    c.age i.sex i.wealth_quintile
```

For binary outcomes, use an appropriate survey-aware model rather than an ordinary model with weights added informally.

## A practical reporting standard

For descriptive MICS estimates, we recommend reporting enough information for readers to understand precision:

```text
estimate
95% confidence interval
unweighted analytic N
```

For detailed technical work, also retain:

```text
standard error
number of PSUs
number of strata
design effect / CV where useful
```

The unweighted `N` is useful context but should never be used as a substitute for the design-based uncertainty estimate.
