## Prepare the analysis data {#c06-prepare-analysis}

All of the worked examples start from one child-level analysis file.

For the UNICEF-based workflow, combine:

```text
mics6-fs-harmonised
        +
mics6-reading-harmonised
        +
survey design variables
        +
household wealth (for the wealth example)
```

### 1. Read the harmonised files

#### R

```{r eval=FALSE}
library(tidyverse)
library(haven)
library(survey)
library(srvyr)

fs <- read_dta(
  "Data/AFLEARN Harmonised Data/mics6-fs-harmonised.dta"
)

reading <- read_dta(
  "Data/AFLEARN Harmonised Data/mics6-reading-harmonised.dta"
)
```

Merge the reading variables:

```{r eval=FALSE}
d <- fs %>%
  left_join(
    reading,
    by = c(
      "country_iso3",
      "year",
      "HH1",
      "HH2",
      "LN"
    ),
    relationship = "one-to-one",
    suffix = c("", "_reading")
  )
```

#### Stata

```stata
use "Data/AFLEARN Harmonised Data/mics6-fs-harmonised.dta", clear

merge 1:1 country_iso3 year HH1 HH2 LN using ///
    "Data/AFLEARN Harmonised Data/mics6-reading-harmonised.dta"

tab _merge
drop _merge
```

### 2. Retain `PSU` and `stratum` in the AFLEARN FS harmonisation

The current AFLEARN `harmonize-mics-fs-v1.0.R` script retains `fsweight` but does not yet retain `PSU` and `stratum`.

These variables should be added to the harmonised FS output before publishing the worked examples.

They are standard MICS survey-design variables:

```text
PSU       primary sampling unit
stratum   sampling stratum
fsweight  selected-child sampling weight
```

After that small change, every worked example can use the same design specification.

::: {.warningbox-header}
**Do not replace this step by clustering on `HH1` without checking**
:::

::: {.warningbox}
Use the sampling-error variables supplied in the MICS data.
:::

### 3. Create identifiers that are unique across surveys

Because PSU and stratum codes can repeat across countries and years, create pooled identifiers.

#### R

```{r eval=FALSE}
d <- d %>%
  mutate(
    survey_id = paste(country_iso3, year, sep = "_"),
    psu_pool = interaction(
      survey_id, PSU,
      drop = TRUE
    ),
    stratum_pool = interaction(
      survey_id, stratum,
      drop = TRUE
    )
  )
```

#### Stata

```stata
egen survey_id = group(country_iso3 year)
egen psu_pool = group(survey_id PSU)
egen stratum_pool = group(survey_id stratum)
```

### 4. Create the survey design

#### R

```{r eval=FALSE}
des <- d %>%
  as_survey_design(
    ids = psu_pool,
    strata = stratum_pool,
    weights = fsweight,
    nest = TRUE
  )
```

#### Stata

```stata
svyset psu_pool [pw=fsweight], strata(stratum_pool)
```

### 5. Construct common numeracy measures

The AFLEARN FS file currently retains the harmonised item responses but does not derive numeracy scores.

The standard harmonised names are:

```text
number_id_*        6 number-identification items
number_compare_*   5 number-discrimination items
number_add_*       5 addition items
number_pattern_*   5 number-pattern items
```

Responses are coded so that:

```text
1 = correct
2 = incorrect
3 = no attempt
```

The standard questionnaire also contains stop rules. For a descriptive total-correct score, this guide scores an administered item as 1 if correct and 0 otherwise, and scores items not reached because of the numeracy stop rule as 0 for children who participated in the FLS assessment.

This produces a score from 0 to 21.

#### R helper

```{r eval=FALSE}
score_correct <- function(x) {
  case_when(
    x == 1 ~ 1,
    x %in% c(2, 3) ~ 0,
    TRUE ~ NA_real_
  )
}

num_items <- c(
  grep("^number_id_", names(d), value = TRUE),
  grep("^number_compare_", names(d), value = TRUE),
  grep("^number_add_", names(d), value = TRUE),
  grep("^number_pattern_", names(d), value = TRUE)
)

length(num_items)
# should be 21
```

Create item scores:

```{r eval=FALSE}
d <- d %>%
  mutate(
    across(
      all_of(num_items),
      score_correct,
      .names = "{.col}_score"
    )
  )
```

For children with a completed/usable FLS assessment, derive:

```{r eval=FALSE}
score_vars <- paste0(num_items, "_score")

d <- d %>%
  rowwise() %>%
  mutate(
    numeracy_total = if_else(
      age >= 7 & age <= 14 & fl_child_result == 1,
      sum(c_across(all_of(score_vars)), na.rm = TRUE),
      NA_real_
    )
  ) %>%
  ungroup()
```

Before publication, AFLEARN should validate this total-score construction against the questionnaire routing in each supported survey.

### 6. Construct task-level measures

For descriptive analysis, calculate the number correct in each domain:

```{r eval=FALSE}
d <- d %>%
  rowwise() %>%
  mutate(
    n_number_id = sum(
      c_across(starts_with("number_id_") & ends_with("_score")),
      na.rm = TRUE
    ),
    n_number_compare = sum(
      c_across(starts_with("number_compare_") & ends_with("_score")),
      na.rm = TRUE
    ),
    n_number_add = sum(
      c_across(starts_with("number_add_") & ends_with("_score")),
      na.rm = TRUE
    ),
    n_number_pattern = sum(
      c_across(starts_with("number_pattern_") & ends_with("_score")),
      na.rm = TRUE
    )
  ) %>%
  ungroup()
```

The final production version should also include a validated `foundational_numeracy` variable reproducing the published MICS indicator.

### 7. Check everything before analysis

At minimum:

```{r eval=FALSE}
d %>%
  count(country_iso3, year)

summary(d$fsweight)

d %>%
  summarise(
    missing_psu = sum(is.na(PSU)),
    missing_stratum = sum(is.na(stratum))
  )

table(d$reading_status, useNA = "ifany")
summary(d$numeracy_total)
```

The worked examples assume these checks have passed.
