## Who is assessed? {#c06-who-assessed}

### Research question

> Who is represented in the MICS6 selected-child sample, and who is eligible for the Foundational Learning Skills assessment?

This is a useful first analysis because the `fs` file contains one randomly selected child aged 5–17 per household, while the FLS assessment targets ages 7–14.

### What to examine

For each survey, describe:

- age distribution of the `fs` sample;
- proportion aged 7–14;
- current school attendance by age;
- grade among currently enrolled children; and
- number of children contributing to learning outcomes.

### Weighted age distribution

#### R

```{r eval=FALSE}
des %>%
  group_by(country_iso3, age) %>%
  summarise(
    share = survey_mean(
      vartype = "ci",
      proportion = TRUE
    )
  )
```

For one country:

```{r eval=FALSE}
country_des <- des %>%
  filter(country_iso3 == "MWI")

country_des %>%
  group_by(age) %>%
  summarise(
    share = survey_mean(proportion = TRUE)
  )
```

#### Stata

```stata
svy: proportion age if country_iso3 == "MWI"
```

### School attendance by age

The harmonised variable `enrolled` comes from `CB7`.

In standard MICS coding:

```text
1 = yes
2 = no
```

#### R

```{r eval=FALSE}
des %>%
  filter(age >= 7, age <= 14) %>%
  mutate(
    in_school = enrolled == 1
  ) %>%
  group_by(country_iso3, age) %>%
  summarise(
    enrolled_share = survey_mean(
      in_school,
      vartype = "ci",
      na.rm = TRUE
    )
  )
```

#### Stata

```stata
gen in_school = enrolled == 1 if enrolled < .

svy, subpop(if inrange(age,7,14)): ///
    mean in_school, over(country_iso3 age)
```

### A useful first figure

Plot school attendance by age for each country.

```text
x-axis: age 7–14
y-axis: proportion currently attending
line: country
```

This immediately shows how much the composition of the assessed population differs across countries as children get older.

### Interpretation

The important point is not simply whether school attendance is high or low.

MICS allows researchers to see that:

- the learning assessment population is defined by age and household sampling, not school enrolment;
- out-of-school children remain part of the target population; and
- the composition of children in a particular grade can differ substantially across countries.

This is one reason age-based and grade-based learning estimates answer different questions.

### Output to add after running

Add:

1. a figure of current attendance by age;
2. a small table showing the unweighted `fs` N and FLS-age N by survey; and
3. one paragraph noting the main cross-country patterns.
