# Worked example 2: Indicators versus distributions {#c06-numeracy-distributions}

## Research question

> Does the headline foundational-numeracy indicator tell the same story as the underlying distribution of numeracy performance?

MICS6 contains 21 numeracy items across four domains:

- 6 number-identification items;
- 5 number-discrimination items;
- 5 addition items; and
- 5 number-pattern items.

A binary indicator is useful for communication, but it compresses a large amount of information.

## Construct the total score

Using the preparation code, define:

```text
numeracy_total = number of items correct out of 21
```

The production version should also contain:

```text
foundational_numeracy
```

validated against published MICS tabulations.

## Compare the two measures

### R

```{r eval=FALSE}
num_des <- des %>%
  filter(age >= 7, age <= 14)

num_des %>%
  group_by(country_iso3) %>%
  summarise(
    mean_score = survey_mean(
      numeracy_total,
      vartype = "ci",
      na.rm = TRUE
    ),
    foundational = survey_mean(
      foundational_numeracy,
      vartype = "ci",
      na.rm = TRUE
    )
  )
```

### Stata

```stata
svy, subpop(if inrange(age,7,14)): ///
    mean numeracy_total foundational_numeracy, ///
    over(country_iso3)
```

## Look at the score distribution

For one country:

```{r eval=FALSE}
des %>%
  filter(
    country_iso3 == "MWI",
    age >= 7,
    age <= 14
  ) %>%
  group_by(numeracy_total) %>%
  summarise(
    share = survey_mean(
      proportion = TRUE,
      na.rm = TRUE
    )
  )
```

A histogram or weighted frequency plot can make the distribution visible.

## Compare the four domains

```{r eval=FALSE}
des %>%
  filter(age >= 7, age <= 14) %>%
  group_by(country_iso3) %>%
  summarise(
    number_id = survey_mean(
      n_number_id / 6,
      na.rm = TRUE
    ),
    compare = survey_mean(
      n_number_compare / 5,
      na.rm = TRUE
    ),
    addition = survey_mean(
      n_number_add / 5,
      na.rm = TRUE
    ),
    pattern = survey_mean(
      n_number_pattern / 5,
      na.rm = TRUE
    )
  )
```

These are **average shares of items correct**, not the official MICS task indicators. Label them accordingly.

## Why this matters

Consider two children:

```text
Child A: 20/21
Child B:  4/21
```

A stringent binary proficiency measure can classify both below the benchmark even though their observed skill profiles are very different.

This matters particularly when:

- overall proficiency is low;
- comparing socioeconomic groups;
- studying progress by grade; or
- diagnosing which components of numeracy are most difficult.

## Suggested figure

A useful two-panel publication figure would show:

```text
Panel A: % meeting foundational numeracy
Panel B: mean total score out of 21
```

for the same countries.

The comparison makes visible where the ranking or apparent size of cross-country differences changes when the full score distribution is used.

## Output to add after running

Add:

- country means and confidence intervals;
- the proficiency estimates;
- one example score distribution; and
- a brief discussion of where the binary indicator compresses meaningful variation.
