# Worked example 3: Why is the reading outcome missing? {#c06-reading-status}

## Research question

> When a child has no reading score, what happened in the assessment?

A missing reading score does not have one meaning.

AFLEARN's `reading_status` reconstructs the reading pathway.

The current coding is:

| Code | Meaning |
|---:|---|
| 0 | Not interviewed |
| 1 | Outside FLS reading age |
| 2 | Caregiver consent not given |
| 3 | Child consent not given |
| 4 | Assessment-language mismatch |
| 5 | Child did not want to read the story |
| 6 | Failed the practice sentence/questions |
| 7 | Attempted a reading passage |
| 8 | Met the foundational-reading criterion |
| 9 | Not classified / inconsistent routing |

Codes 7 and 8 both indicate that the child reached the passage; code 8 additionally identifies children meeting the foundational-reading criterion.

## Describe the pathway among ages 7–14

### R

```{r eval=FALSE}
reading_status_tab <- des %>%
  filter(age >= 7, age <= 14) %>%
  group_by(country_iso3, reading_status) %>%
  summarise(
    share = survey_mean(
      proportion = TRUE,
      vartype = "ci",
      na.rm = TRUE
    )
  )
```

### Stata

```stata
svy, subpop(if inrange(age,7,14)): ///
    proportion reading_status, over(country_iso3)
```

## A simpler attempted/not-attempted distinction

```{r eval=FALSE}
d <- d %>%
  mutate(
    reading_attempted = reading_status %in% c(7, 8)
  )
```

Then:

```{r eval=FALSE}
des <- d %>%
  as_survey_design(
    ids = psu_pool,
    strata = stratum_pool,
    weights = fsweight,
    nest = TRUE
  )

des %>%
  filter(age >= 7, age <= 14) %>%
  group_by(country_iso3) %>%
  summarise(
    attempted = survey_mean(
      reading_attempted,
      vartype = "ci",
      na.rm = TRUE
    )
  )
```

## Why not code all non-attempts as zero reading skill?

Because these categories describe different processes.

For example:

```text
language mismatch
```

means the survey could not offer an appropriate reading passage under the questionnaire routing.

That is not direct evidence that the child cannot read.

Similarly, lack of caregiver consent is a nonresponse event, not an observed low score.

By contrast, failing the practice sequence is much closer to evidence about the child's reading performance.

The appropriate treatment depends on the research question.

## Recommended reporting

Whenever reading results exclude a substantial share of age-eligible children, report:

```text
% of target-age children with an attempted passage
```

alongside the learning outcome.

This makes assessment coverage visible rather than allowing it to disappear into missing values.

## Suggested figure

A stacked bar by country:

```text
attempted passage
practice-route non-attempt
language mismatch
consent/refusal
other
```

Keep categories substantively meaningful rather than displaying all ten numeric codes if several are rare.

## Output to add after running

Add one figure and identify countries where assessment coverage is materially lower than the nominal 7–14 target population.
