# Worked example 4: Reading language and assessment coverage {#c06-reading-language}

## Research question

> In which language was each child assessed, and how does assessment-language coverage vary?

Reading is different from numeracy because it requires a language-specific passage.

The harmonised reading file contains:

```text
passage_language
```

and AFLEARN's routing variable identifies:

```text
reading_status == 4
```

for an assessment-language mismatch.

## Distribution of assessment languages

Among children who attempted a passage:

### R

```{r eval=FALSE}
des %>%
  filter(
    age >= 7,
    age <= 14,
    reading_status %in% c(7, 8)
  ) %>%
  group_by(country_iso3, passage_language) %>%
  summarise(
    share = survey_mean(
      proportion = TRUE,
      na.rm = TRUE
    )
  )
```

### Stata

```stata
gen read_attempt = inlist(reading_status,7,8)

svy, subpop(if inrange(age,7,14) & read_attempt): ///
    proportion passage_language, over(country_iso3)
```

## Language mismatch among all age-eligible children

### R

```{r eval=FALSE}
des %>%
  filter(age >= 7, age <= 14) %>%
  mutate(
    language_mismatch = reading_status == 4
  ) %>%
  group_by(country_iso3) %>%
  summarise(
    mismatch = survey_mean(
      language_mismatch,
      vartype = "ci",
      na.rm = TRUE
    )
  )
```

## Relate assessment language to home and school language

The FS harmonised file contains:

```text
lang_home
lang_school
```

but these codes are **country-specific**.

Do not pool their numeric values across countries.

Within one survey, however, you can examine alignment.

For example:

```{r eval=FALSE}
d %>%
  filter(country_iso3 == "MWI") %>%
  count(
    lang_home,
    lang_school,
    passage_language
  )
```

For weighted percentages, use the survey design object.

## Malawi as a validation example

Malawi is especially useful because the survey administered English and Chichewa passages.

After merging the AFLEARN reading file, both passage languages should appear among assessed children.

This is an important check because the IPUMS reading variables retain only one passage in this multi-passage survey.

## Interpretation

A lower reading-assessment rate can arise partly from the relationship between:

- home language;
- school language;
- available assessment language; and
- questionnaire routing.

This is not merely a technical issue.

If children from particular linguistic groups are less likely to receive a valid reading assessment, the observed reading sample differs from the target-age population.

## Suggested outputs

For multilingual surveys, show:

1. share assessed in each passage language; and
2. share of age-eligible children with assessment-language mismatch.

For monolingual/single-passage surveys, a table may be sufficient.
