## Learning inequalities by wealth {#c06-wealth}

### Research question

> Do wealth inequalities look the same when learning is measured using a binary proficiency indicator versus the full score distribution?

This example illustrates why outcome choice matters.

When overall proficiency is very low, children in both the poorest and richest quintiles may mostly fall below the benchmark. A binary indicator can therefore make the groups look more similar than their underlying performance actually is.

### Data requirement

The current AFLEARN FS harmonised file does not contain wealth.

Add `windex5` from the household file, or request `WINDEX5` in an IPUMS extract.

Standard MICS wealth quintiles are:

```text
1 poorest
2 second
3 middle
4 fourth
5 richest
```

### Compare proficiency by wealth

#### R

```{r eval=FALSE}
des %>%
  filter(age >= 7, age <= 14) %>%
  group_by(country_iso3, windex5) %>%
  summarise(
    foundational_num = survey_mean(
      foundational_numeracy,
      vartype = "ci",
      na.rm = TRUE
    )
  )
```

#### Stata

```stata
svy, subpop(if inrange(age,7,14)): ///
    mean foundational_numeracy, ///
    over(country_iso3 windex5)
```

### Compare average numeracy score by wealth

#### R

```{r eval=FALSE}
des %>%
  filter(age >= 7, age <= 14) %>%
  group_by(country_iso3, windex5) %>%
  summarise(
    mean_num = survey_mean(
      numeracy_total,
      vartype = "ci",
      na.rm = TRUE
    )
  )
```

#### Stata

```stata
svy, subpop(if inrange(age,7,14)): ///
    mean numeracy_total, ///
    over(country_iso3 windex5)
```

### Focus on poorest versus richest

A compact inequality measure is the richest-poorest difference.

In R, first estimate each quintile and then calculate the contrast, or use a survey regression.

```{r eval=FALSE}
model <- svyglm(
  numeracy_total ~ factor(windex5),
  design = subset(
    des,
    age >= 7 & age <= 14
  )
)

summary(model)
```

The coefficient on the richest category relative to the poorest gives the adjusted-by-design difference in mean score, with no other covariates.

### Why the two inequality measures can differ

Imagine:

```text
Poorest: mean 5/21, proficiency 1%
Richest: mean 10/21, proficiency 4%
```

The proficiency gap is only three percentage points, but the average score differs by five items.

The benchmark is not wrong. It answers a different question:

> What share reaches a demanding threshold?

The mean score asks:

> How far through the observed skill progression are children?

For inequality analysis, showing both is often more informative.

### Suggested figure

For each country, plot two wealth gradients:

```text
Panel A: proficiency by wealth quintile
Panel B: mean score by wealth quintile
```

Use the same quintile ordering and clearly different y-axis labels.

### Extension to reading

The same logic can be applied to `max_reading_skills`, `max_reading_accuracy` and `max_reading_score`, provided the reading analysis sample and `reading_status` are handled explicitly.

### Output to add after running

Select two or three countries that illustrate different patterns rather than reproducing all countries in the body of the guide. The full results can remain in downloadable code/output.
