## In-school and out-of-school children {#c06-schooling-status}

### Research question

> At the same age, how does learning differ between children who are currently in school and those who are out of school?

This is one of the distinctive advantages of a household-based learning assessment.

A school-based assessment cannot observe children who are not enrolled or not attending the sampled schools. MICS can.

### Define current schooling status

The harmonised variable `enrolled` records whether the child attended school or an early childhood education programme during the current school year.

For school-age analysis:

```{r eval=FALSE}
d <- d %>%
  mutate(
    in_school = case_when(
      enrolled == 1 ~ 1,
      enrolled == 2 ~ 0,
      TRUE ~ NA_real_
    )
  )
```

Re-create the design object after adding the variable.

### Why compare within age?

A simple comparison of all in-school and out-of-school children can be misleading because the groups may have very different age distributions.

Older adolescents are more likely to have left school in many settings and have also had more potential exposure to schooling.

Compare children of the same age, or adjust for age.

### Descriptive comparison by age

#### R

```{r eval=FALSE}
des %>%
  filter(
    age >= 10,
    age <= 14
  ) %>%
  group_by(
    country_iso3,
    age,
    in_school
  ) %>%
  summarise(
    mean_num = survey_mean(
      numeracy_total,
      vartype = "ci",
      na.rm = TRUE
    )
  )
```

#### Stata

```stata
svy, subpop(if inrange(age,10,14)): ///
    mean numeracy_total, ///
    over(country_iso3 age in_school)
```

### Age-adjusted comparison

#### R

```{r eval=FALSE}
model <- svyglm(
  numeracy_total ~ factor(age) + factor(in_school),
  design = subset(
    des,
    age >= 10 & age <= 14
  )
)

summary(model)
```

This estimates the association between current school attendance and numeracy conditional on age.

It is not a causal estimate of the effect of attending school.

Children who remain in school differ from children who leave school in many observed and unobserved ways.

### Reading

The same analysis can be performed for reading, but first define the relevant reading population using `reading_status`.

```{r eval=FALSE}
read_des <- des %>%
  filter(
    age >= 10,
    age <= 14,
    reading_status %in% c(7, 8)
  )
```

Then compare `max_reading_accuracy` or another clearly defined reading outcome.

### Interpretation

The value of this analysis is descriptive:

> What learning levels are observed among children who are currently in school and among children who are not, at comparable ages?

Do not write:

> School attendance increases numeracy by X points.

The household-survey data do not identify that causal effect without a separate research design.

### Suggested figure

For ages 10–14:

```text
x-axis: age
y-axis: mean numeracy score
two lines: in school / out of school
```

Facet by country or select illustrative countries.

Also show the number of out-of-school observations contributing to each estimate; some age-country cells may be too small for useful inference.
