# Worked example 7: Learning by grade and age {#c06-grade-trajectories}

## Research question

> How does learning change across grades, and how should we interpret differences when children of different ages occupy the same grade?

Grade trajectories are intuitive, but MICS requires more care than a school-based grade assessment.

## First: construct a meaningful grade variable

The AFLEARN harmonised variable:

```text
current_grade
```

is the year **within the education level**.

It cannot be treated as a universal grade number without `current_level_h`.

For primary school:

```{r eval=FALSE}
d_primary <- d %>%
  filter(
    enrolled == 1,
    current_level_h == 1
  )
```

Within this restricted group, `current_grade` can be interpreted as primary year/grade.

If you want a single grade scale spanning primary and secondary, construct it from the country's education system rather than simply using `current_grade`.

See [Why current_grade is not the child's grade](#c07-current-grade).

## Mean numeracy by primary grade

### R

```{r eval=FALSE}
des %>%
  filter(
    age >= 7,
    age <= 14,
    enrolled == 1,
    current_level_h == 1,
    current_grade >= 1,
    current_grade <= 6
  ) %>%
  group_by(
    country_iso3,
    current_grade
  ) %>%
  summarise(
    mean_num = survey_mean(
      numeracy_total,
      vartype = "ci",
      na.rm = TRUE
    )
  )
```

### Stata

```stata
gen primary = current_level_h == 1

svy, subpop(if inrange(age,7,14) & enrolled==1 & ///
    primary & inrange(current_grade,1,6)): ///
    mean numeracy_total, ///
    over(country_iso3 current_grade)
```

## Now examine age within grade

### R

```{r eval=FALSE}
des %>%
  filter(
    enrolled == 1,
    current_level_h == 1
  ) %>%
  group_by(
    country_iso3,
    current_grade
  ) %>%
  summarise(
    mean_age = survey_mean(
      age,
      vartype = "ci",
      na.rm = TRUE
    )
  )
```

A country with an average age of 10.5 in Grade 3 is not compositionally identical to one with an average age of 8.5 in Grade 3.

Late entry, repetition and interruptions to schooling all affect the relationship between age and grade.

## Age and grade are jointly informative

A useful model is:

```{r eval=FALSE}
model <- svyglm(
  numeracy_total ~ factor(current_grade) + age,
  design = subset(
    des,
    enrolled == 1 &
      current_level_h == 1 &
      age >= 7 &
      age <= 14
  )
)

summary(model)
```

This describes grade differences after accounting linearly for age.

Again, it is descriptive rather than causal.

## Suggested figure

For each country:

```text
x-axis: primary grade
y-axis: mean numeracy score
line: country
```

and accompany it with a second figure/table showing mean age by grade.

This prevents the trajectory graph from being interpreted as if every country's Grade 3 population had the same age composition.

## Cross-country trajectory tests

If the research question is whether the **shape** of the grade trajectory differs across countries, use a regression with country-by-grade interactions rather than eyeballing the plot.

For a flexible quadratic trajectory:

```{r eval=FALSE}
model <- svyglm(
  numeracy_total ~
    factor(country_iso3) *
    (current_grade + I(current_grade^2)),
  design = subset(
    des,
    enrolled == 1 &
      current_level_h == 1 &
      current_grade >= 1 &
      current_grade <= 6
  )
)
```

Jointly test the country-by-grade interaction terms.

In Stata:

```stata
svy: regress numeracy_total ///
    i.country_iso3##c.current_grade##c.current_grade ///
    if enrolled==1 & current_level_h==1 & ///
       inrange(current_grade,1,6)

testparm i.country_iso3#c.current_grade ///
         i.country_iso3#c.current_grade#c.current_grade
```

The interaction test asks whether the trajectories differ across countries, not merely whether average performance levels differ.
