## Timing within the school year {#c06-school-timing}

### Research question

> If two children are both labelled "Grade 3", have they necessarily had the same amount of exposure to Grade 3 instruction?

No.

MICS fieldwork can occur at different points in the school year across countries. A Grade 3 child assessed one month after the school year begins has had less Grade 3 exposure than a Grade 3 child assessed near the end of the year.

This matters when interpreting cross-country grade trajectories.

### What you need

AFLEARN's harmonised FS file already contains `interview_month`, but the current processing does not contain a harmonised variable measuring the child's position within the country's school year.

For this example, create `school_month` where:

```text
0 = beginning of school year
1 = one month after beginning
...
```

This requires a country-level lookup table giving the school-year start month.

### Example lookup

Illustrative structure:

```text
country_iso3   school_start_month
BEN            ...
GHA            ...
MWI            ...
```

The values must be verified from country education calendars or survey documentation before publication.

### Convert interview month to months through school year

#### R

```{r eval=FALSE}
d <- d %>%
  left_join(
    school_calendar,
    by = "country_iso3"
  ) %>%
  mutate(
    school_month = (
      interview_month - school_start_month
    ) %% 12
  )
```

Depending on survey timing and holidays, a simple month difference may need country-specific adjustment.

### Construct grade progress

For children in a comparable grade system:

```{r eval=FALSE}
d <- d %>%
  mutate(
    gradeprogress =
      current_grade + school_month / 12
  )
```

Thus:

```text
Grade 3, start of year       ≈ 3.0
Grade 3, six months in       ≈ 3.5
Grade 4, start of year       ≈ 4.0
```

This is an approximation to instructional exposure, not a literal measure of time in school.

### Fit a trajectory

#### R

```{r eval=FALSE}
model <- svyglm(
  numeracy_total ~
    factor(country_iso3) *
    (gradeprogress + I(gradeprogress^2)),
  design = subset(
    des,
    enrolled == 1 &
      current_level_h == 1 &
      gradeprogress >= 1 &
      gradeprogress <= 6
  )
)

summary(model)
```

Test whether trajectories differ across countries:

```{r eval=FALSE}
library(regTermTest)

regTermTest(
  model,
  ~ factor(country_iso3):gradeprogress +
    factor(country_iso3):I(gradeprogress^2)
)
```

Depending on the fitted model and package behaviour, the exact joint-test syntax may need adjustment. The production script should verify the coefficient names and test matrix explicitly.

#### Stata

```stata
gen gradeprogress = current_grade + school_month/12

svy: regress numeracy_total ///
    i.country_iso3##c.gradeprogress##c.gradeprogress ///
    if enrolled==1 & current_level_h==1 & ///
       inrange(gradeprogress,1,6)

testparm i.country_iso3#c.gradeprogress ///
         i.country_iso3#c.gradeprogress#c.gradeprogress
```

### Predicted trajectories

In Stata:

```stata
margins country_iso3, ///
    at(gradeprogress=(1(0.25)6))

marginsplot
```

In R, generate a prediction grid and use `predict()` with standard errors, or use a survey-compatible marginal-estimation package after verifying its handling of `svyglm`.

### Why this example is more advanced

`gradeprogress` solves only one comparability problem.

It does not account for differences in official school-entry age, grade repetition, interruptions to schooling, variation in instructional time, attendance, curriculum differences, or differences in the meaning of grade across education systems.

It should therefore be described as an adjustment for survey timing within the school year, not a complete measure of schooling exposure.

This is the final, advanced worked example. It demonstrates how the rich timing information in MICS can improve cross-country comparisons while also showing why researchers need to be explicit about the assumptions behind derived variables.
