## Why current_grade is not the child's grade {#c07-current-grade}

`current_grade` looks like a grade variable, but it records the year within the child's current education level.

That distinction is easy to miss.

Consider:

| `current_level_h` | `current_grade` | Meaning |
|---|---:|---|
| Primary | 1 | First year of primary |
| Lower secondary | 1 | First year of lower secondary |
| Upper secondary | 1 | First year of upper secondary |

All three children have `current_grade = 1`, but they are at very different points in the school system.

The same issue applies to `highest_grade` and `previous_grade`, which should be interpreted with their corresponding education-level variables.

### Why this happens

The standard MICS schooling questions separately record education level and grade/year within that level.

AFLEARN harmonises the level into `current_level_h` but deliberately leaves `current_grade` as the year reported within that level.

This preserves the source information rather than imposing a single grade system on countries with different education structures.

### The "sawtooth" problem

If you pool currently enrolled children and graph only `current_grade`, you will see many children at 1, 2, 3 because primary, lower secondary and upper secondary cycles each restart.

That distribution is not a simple Grade 1 → Grade 12 progression.

Before using grade for trajectories, either:

1. restrict the analysis to one education level; or
2. construct a country-specific administrative grade.

### Option 1: restrict to primary school

For many foundational-learning analyses, the cleanest solution is:

```{r eval=FALSE}
primary <- d %>%
  filter(
    enrolled == 1,
    current_level_h == 1
  )
```

Within primary, `current_grade = 1, 2, 3, ...` has a much clearer interpretation.

This is the approach used in the basic grade-trajectory example in [Learning by grade and age](#c06-grade-trajectories).

### Option 2: construct an administrative grade

Suppose a country has:

```text
6 years primary
3 years lower secondary
3 years upper secondary
```

Then:

```{r eval=FALSE}
d <- d %>%
  mutate(
    admin_grade = case_when(
      current_level_h == 1 ~ current_grade,
      current_level_h == 2 ~ 6 + current_grade,
      current_level_h == 3 ~ 9 + current_grade,
      TRUE ~ NA_real_
    )
  )
```

produces:

```text
Primary 1          admin_grade 1
...
Primary 6          admin_grade 6
Lower secondary 1  admin_grade 7
...
Upper secondary 3  admin_grade 12
```

### Ghana example

In the current MICS6 harmonised data, Ghana provides a clean example of a `6 + 3 + 3` structure.

The administrative-grade transformation is therefore:

```{r eval=FALSE}
gha <- d %>%
  filter(country_iso3 == "GHA") %>%
  mutate(
    admin_grade = case_when(
      current_level_h == 1 ~ current_grade,
      current_level_h == 2 ~ 6 + current_grade,
      current_level_h == 3 ~ 9 + current_grade,
      TRUE ~ NA_real_
    )
  )
```

### Do not use the same offsets in every country

Primary-cycle length differs across countries.

The current source data include examples of:

```text
5-year primary cycles
6-year primary cycles
7-year primary cycles
8-year primary cycles
```

Some systems also code lower-secondary years differently or do not use a separate lower-secondary category in the same way.

A correct administrative-grade variable therefore requires a country-specific education-system lookup.

A useful future AFLEARN lookup would contain:

```text
country_iso3
survey_year
current_level_h
grade_within_level
admin_grade
notes
source
```

### Validate the constructed grade

After constructing `admin_grade`, check:

1. plausible ages by grade;
2. weighted enrolment distribution;
3. maximum grade against the education structure;
4. transitions between education levels; and
5. country documentation.

For example:

```{r eval=FALSE}
des %>%
  filter(enrolled == 1) %>%
  group_by(country_iso3, admin_grade) %>%
  summarise(
    mean_age = survey_mean(
      age,
      na.rm = TRUE
    )
  )
```

A wildly non-monotonic age-by-grade profile is a sign that the grade construction needs review.

### Grade is still not schooling exposure

Even a correctly constructed administrative grade does not capture repetition, late school entry, interrupted schooling, attendance, time through the current school year, or instructional time.

For cross-country learning trajectories, [Timing within the school year](#c06-school-timing) therefore also introduces `gradeprogress`, which adjusts grade approximately for the timing of the MICS interview within the school year.

That variable answers a different problem and requires its own documented assumptions.
