# Validation and reproducibility {#c07-validation}

A harmonisation script can run without errors and still produce the wrong data.

AFLEARN therefore recommends validating at three levels:

```text
structural checks
      +
distribution checks
      +
external checks
```

## 1. Structural checks

For every survey, verify:

### Unique selected-child identifiers

```{r eval=FALSE}
d %>%
  count(HH1, HH2, LN) %>%
  filter(n > 1)
```

Expected:

```text
0 rows
```

### Expected survey size

Compare the number of harmonised FS records with the original `fs.sav`.

A harmonisation that unexpectedly drops children should fail loudly or report the exclusions explicitly.

### Required fields

Check that required variables exist:

```{r eval=FALSE}
required <- c(
  "country_iso3",
  "year",
  "HH1",
  "HH2",
  "LN",
  "age",
  "fsweight"
)

setdiff(required, names(d))
```

For the revised guide also require:

```text
PSU
stratum
```

## 2. Range and coding checks

Examples:

```{r eval=FALSE}
range(d$age, na.rm = TRUE)

table(
  d$current_level_h,
  useNA = "ifany"
)

summary(d$fsweight)

table(
  d$reading_status,
  useNA = "ifany"
)
```

For reading:

```{r eval=FALSE}
summary(reading$reading_accuracy)

reading %>%
  filter(reading_accuracy > 1)
```

Any accuracy above 1 requires investigation.

For numeracy, after creating a total score:

```{r eval=FALSE}
range(d$numeracy_total, na.rm = TRUE)
```

should remain within:

```text
0 to 21
```

## 3. Routing checks

Variables should follow the questionnaire logic.

Examples:

```text
children outside ages 7–14
→ should not have a normal FLS reading outcome

failed practice
→ should not have a valid main-passage score unless the country questionnaire legitimately permits it

reading_status = attempted
→ should have a valid passage/score

reading_status = foundational
→ should meet the underlying word/comprehension criteria
```

Write these as automated tests where possible.

For example:

```{r eval=FALSE}
stopifnot(
  all(
    is.na(reading$reading_score[
      reading$reading_status == 1
    ])
  )
)
```

Country exceptions should be documented rather than weakening the test globally.

## 4. Cross-file merge checks

Whenever `hl`, `hh`, reading or IPUMS data are linked:

- confirm key uniqueness before the merge;
- report unmatched records;
- confirm the number of analysis rows after the merge; and
- compare key distributions before and after.

Do not routinely discard `_merge != 3` or unmatched rows without understanding them.

## 5. Compare with published MICS results

For indicators that MICS itself publishes, reproduction is one of the strongest validation checks available.

Examples include:

- foundational-reading classifications;
- foundational numeracy;
- school attendance;
- selected household/child characteristics.

A discrepancy should trigger investigation of:

```text
denominator
age restriction
weight
routing
indicator definition
rounding convention
survey year
```

rather than an immediate assumption that either source is wrong.

## Reading-threshold validation

AFLEARN's rounded-down integer cutoff for the 90% reading-accuracy requirement was retained because alternative implementations were compared with published MICS classifications and the rounded-down convention reproduced those classifications.

This should remain part of the regression test for the reading script.

If a future edit changes the reading construction, rerun the country-level comparison against the MICS publications.

## 6. Validate multilingual reading separately

For every multi-passage survey check:

```text
number attempting each passage
passage-language distribution
passage length
multiple-attempt count
reading-status distribution
```

For Malawi, the validation should specifically confirm that Chichewa assessments are retained rather than disappearing from the analysis sample.

## 7. Survey-design checks

Once `PSU` and `stratum` are retained in the FS harmonisation, check:

```{r eval=FALSE}
d %>%
  summarise(
    missing_psu = sum(is.na(PSU)),
    missing_stratum = sum(is.na(stratum))
  )
```

and:

```{r eval=FALSE}
d %>%
  distinct(
    country_iso3,
    year,
    stratum,
    PSU
  ) %>%
  count(
    country_iso3,
    year,
    stratum,
    name = "n_psu"
  )
```

Unexpected single-PSU strata or missing design variables should be investigated against the country sampling documentation.

## 8. Version the scripts and outputs together

A published harmonised dataset should identify the exact processing version used.

For example:

```text
FS harmonisation      v1.1
Reading harmonisation v1.2
Guide                  2026-08
```

A useful release structure is:

```text
release/
├── scripts/
├── crosswalks/
├── country-notes/
├── validation/
└── README
```

The validation directory can contain machine-readable summaries such as:

```text
survey_counts.csv
reading_status_checks.csv
published_indicator_comparison.csv
merge_checks.csv
```

## 9. Keep raw data immutable

The processing workflow should never edit the original UNICEF source files.

A reproducible project keeps:

```text
raw source data
      ↓
processing scripts
      ↓
derived harmonised files
```

so every result can be regenerated from the original inputs.

## 10. Minimum validation before adding a new survey

Before a survey is added to the AFLEARN supported list:

- [ ] FS identifiers are unique.
- [ ] Key questionnaire variables have been checked against the survey questionnaire.
- [ ] Education-level mappings have been reviewed.
- [ ] Survey weights/design variables are present.
- [ ] Reading passage(s) and languages have been identified.
- [ ] Passage length has been validated.
- [ ] Reading routing has been checked.
- [ ] Reading classifications reproduce published results where available.
- [ ] Numeracy indicator reproduces published results if constructed.
- [ ] Crosswalk and country notes have been updated.
- [ ] No unexplained records fall into `reading_status = 9`.
- [ ] All output files regenerate cleanly from the scripts.

This makes extending the guide slower than simply appending another country — but substantially safer.
